An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2022-07-12 simplified_sdc_intro/paper.tex
first pass at referee

%\documentclass[times,modern]{aastex63}
\documentclass[times,preprint]{aastex631}
%\documentclass{aastex63}

@@

Modeling astrophysical reacting flows can be challenging because of
the disparity between the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to breakdown.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium. 
These large changes in abundances (and large alternately positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015,rivas:2022}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

@@

In the Strang splitting~\citep{strang:1968} flavor of operator splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  
%A variation on Strang splitting called (re-)balanced
%splitting was developed in \citet{speth:2013}, but we do not explore
%that here. \MarginPar{keep this sentence? MK: Don't think it's needed.}

@@

We begin with the general form of a multi-implicit SDC correction equation with two physical processes, as originally presented in \cite{bourlioux2003high}.
Using $(k)$ to denote the iterate, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
In the spirit of \cite{SDC-old}, we treat each advection term as piecewise constant over the time step,
and equal to a time-centered advection term evaluated with the Godunov integrator.
We note that this is a departure from the approach in \cite{bourlioux2003high} that uses a method of lines approach, where 
the advection terms are computed at each node, rather than the midpoint between nodes, using spatial extrapolation only.
Using $\boldsymbol{\mathcal{A}}(\Uc^{n+1/2})$ to denote the advection term, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
Also following \cite{SDC-old}, we differentiate this equation in time
\begin{eqnarray}
\frac{\partial\Uc^{(k)}}{\partial t} = && \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right]\nonumber\\
&+& \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right].
\end{eqnarray}
We combine terms, and arrive at a system of ODEs at each cell,
\begin{equation}
\frac{\partial\Uc^{(k)}}{\partial t} =  \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) \right].\label{eq:sdcupdate}
\end{equation}
We integrate this system from $t^n$ to $t^{n+1}$ using a modified version
of the VODE~\citep{vode} integrator, as described in Section \ref{sec:VODE}.
Thus, we are effectively using two temporal nodes ($t^n$ and $t^{n+1}$) 
and the overall integration is second-order accurate.
In order to properly couple reactions into the computation of the advection terms, 
we use an iteratively lagged reaction source term,
denoted $\Ic_q$ (this is basically an approximation of $\Rb(\qb)$).
The overall flow of the simplified-SDC algorithm is shown in Figure~\ref{fig:sdc_flowchart}.

@@

    \item Predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU PPM
      method.  The source terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sc(\qb) = \Gb(\qb) + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      $\Ic_\qb^{n+1/2,(k-1)}$ is a numerical approximation for the effect of
      reactions from the previous iterate, which is computed at the end of each iterate 
      as described below.
      The use of this term contrasts with Strang-splitting, where no
      reactive source terms are included in the hydrodynamics update.
      If $k=1$, we instead use the value from the last iteration of the previous
      time step, $\Ic^{n-1/2,(\kmax)}_{\qb}$.
%%       Any hydrodynamic source terms are time-centered
%%       using the previous iteration:
%% \MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
%%       \begin{equation}
%%         \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
%%       \end{equation}

@@

    Our integrator also needs the Jacobian of the system, in terms of
    $\Uc^\prime$.  This is different than the form of the
    Jacobian usually used in reaction networks.  We describe the form
    of the Jacobian in Appendix \ref{sec:app:jacobian}.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported to C++, as a
header-only library enabling it to run on GPUs.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/AMReX-Astro/Microphysics}.}.  We have found empirically that these checks improve the performance of VODE.

@@

Tables \ref{table:react_converge_strang} and \ref{table:react_converge_sdc} show
the Strang splitting and simplified-SDC convergence respectively.  The convergence rate 
is computed by taking log-base-2 of the ratio of adjacent errors.  A rate of 2.0
indicates second order convergence.  We see that both methods converge nearly
second order, and agree well with each other.  This demonstrates that both
methods are working as designed.

>>>>

%\documentclass[times,modern]{aastex63}
\documentclass[linenumbers,trackchanges]{aastex631}
%\documentclass{aastex63}

@@

Modeling astrophysical reacting flows can be challenging because of
the disparity between the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to \replaced{breakdown}{break down, since the reactions do not see the effects of the hydrodynamics and vice versa}.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium. 
These large changes in abundances (and large alternately positive and negative energy generation rates) can be a challenge for
\replaced{a code}{the integration method---it may take more steps than allowed, require a timestep below floating point accuracy, or fail to meet the tolerenaces}.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015,rivas:2022}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

@@

In the Strang splitting~\citep{strang:1968} flavor of operator
splitting, we first integrate the system with reactions terms only (no
advection) over $\Delta t/2$, then integrate the advection terms only
(no reactions) over $\Delta t$, and finally integrate the reaction
terms only over $\Delta t/2$.  \added{The staggering of the reactive update
means the hydro implicitly sees a time-centgered reactive source, making
the update second-order accurate in time.}  This flow is illustrated in
Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation \replaced{gets}{achieved} second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  
%A variation on Strang splitting called (re-)balanced
%splitting was developed in \citet{speth:2013}, but we do not explore
%that here. \MarginPar{keep this sentence? MK: Don't think it's needed.}

@@

We begin with the general form of a multi-implicit SDC correction equation with two physical processes\deleted{, as originally presented in} \citep{bourlioux2003high}.
Using $(k)$ to denote the iterate, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
In the spirit of \cite{SDC-old}, we treat each advection term as piecewise constant over the time step,
and equal to a time-centered advection term evaluated with the Godunov integrator.
We note that this is a departure from the approach in \cite{bourlioux2003high} that uses a method of lines approach, where 
the advection terms are computed at each node, rather than the midpoint between nodes, using spatial extrapolation only.
Using $\boldsymbol{\mathcal{A}}(\Uc^{n+1/2})$ to denote the advection term, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
Also following \cite{SDC-old}, we differentiate this equation in time
\begin{eqnarray}
\frac{\partial\Uc^{(k)}}{\partial t} = && \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right]\nonumber\\
&+& \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right].
\end{eqnarray}
We combine terms, and arrive at a system of ODEs at each cell,
\begin{equation}
\frac{\partial\Uc^{(k)}}{\partial t} =  \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) \right].\label{eq:sdcupdate}
\end{equation}
We integrate this system from $t^n$ to $t^{n+1}$ using a modified version
of the VODE~\citep{vode} integrator, as described in Section \ref{sec:VODE}.
Thus, we are effectively using two temporal nodes ($t^n$ and $t^{n+1}$) 
and the overall integration is second-order accurate.
In order to properly couple reactions into the computation of the advection terms, 
we use an iteratively lagged reaction source term,
denoted $\Ic_q$ (this is basically an approximation of $\Rb(\qb)$).
The overall flow of the simplified-SDC algorithm is shown in Figure~\ref{fig:sdc_flowchart}.

@@

    \item Predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU
      PPM method\added{---this involves taking the cell-centered
        primitive state and tracking all of the information that can
        reach the interface over a timestep, following the procedure
        from \citet{ppmunsplit,millercolella:2002}}.  The source
      terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sc(\qb) = \Gb(\qb) + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      $\Ic_\qb^{n+1/2,(k-1)}$ is a numerical approximation for the effect of
      reactions from the previous iterate, which is computed at the end of each iterate 
      as described below.
      The use of this term contrasts with Strang-splitting, where no
      reactive source terms are included in the hydrodynamics update.
      If $k=1$, we instead use the value from the last iteration of the previous
      time step, $\Ic^{n-1/2,(\kmax)}_{\qb}$.
%%       Any hydrodynamic source terms are time-centered
%%       using the previous iteration:
%% \MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
%%       \begin{equation}
%%         \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
%%       \end{equation}

@@

    Our integrator also needs the Jacobian of the system, in terms of
    $\Uc^\prime$.  This is different than the form of the Jacobian
    usually used in reaction networks \added{(we depend on $e$ instead of
    $T$)}.  We describe the form of the Jacobian in Appendix
    \ref{sec:app:jacobian}.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported to \replaced{C++, as a
header-only library enabling it to run on GPUs}{modern C++ and designed to run on GPUs}.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/AMReX-Astro/Microphysics}.}.  We have found empirically that these checks improve the performance of VODE.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang splitting and
simplified-SDC convergence respectively.  \added{We ran each integration method on a domain with $64^2$, $128^2$, $256^2$, $512^2$ zones.  An error is defined between successive resolutions by coarsening the higher resolution run and taking the L1 norm of the zone-by-zone difference of the two simulations---for example, $\epsilon_{64\rightarrow 128}$ is the error between the $64^2$ and $128^2$ simulations.}  \deleted{The convergence rate is
computed by taking log-base-2 of the ratio of adjacent errors.}  \added{The convergence
rate from $\epsilon_{64\rightarrow 128}$ to $\epsilon_{128\rightarrow 256}$ is computed as $\log_2 \epsilon_{64\rightarrow 128}/\epsilon_{128\rightarrow 256}$ and is given in the column between those errors, and likewise for the higher-resolution case.}  A rate
of 2.0 indicates second order convergence.  We see that both methods
\replaced{converge nearly second order, and agree well with each
  other}{exhibit convergence at roughly second-order and agree well.}.
This demonstrates that both methods are working as designed.
